---
title: "LaCombe_Rep"
output: pdf_document
date: "2024-09-26"
---
## Introduction

This tutorial will walk through basic methodology of testing for assortative mixing on a network dataset.

Two R packages are reviewed for this approach: Igraph and assortnet.
Assortative Mixing can occur in two distinct ways. Homophily describes a network in which we see that similar nodes tend to be more connected throughout the network; "birds of a feather flock together." Heterophily describes node behavior in a network in which non-similar nodes tend to be more connected. This tutorial will provide examples of each and how to detect for both.

Another consideration before running code is what exactly the "thing," or more technically the variable, that we want to test. Describing the variable is as simple as using basic statistical terms. We either have a categorical variable (race, gender, landtype, PartyID) that we can identify and measure for each node. Or we have a continuous, numeric values (income, time, height, years of residency). Newman (2003) discusses nominal variables as "discrete" variables, and continuous as "scalar," which matches the technical languages of matrix algebra. I spend time to discuss these differences because it is necessary to understand what attribute in your network you are trying to test, as the commands do depend on this attribute. 

You will also want to consider whether your network is directed. For continuous/scalar attributes, the direction of the ties could matter in your network. This is easily handled by the packages, but do note the part of the code where you should enter if the network is directed. 

Please select the "outline" button at the top right corner of this window to easily skip around in the file to the section that contains the materials appropriate for your assortative network analysis task.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#this is a for loop to install and load all the packages you will need. Reference the list to see what we use.
#Packages
P<-c("network", "assortnet", "readr", "haven", "usmap", "devtools", "ggplot2", "ggeasy")

for (i in 1:length(P)) {
  ifelse(!require(P[i],character.only=TRUE),install.packages(P[i]), #You may have to run this code chunk a few times before it turns green and prints the stars of succesfull loading.
         print("✩")) #Stars will print if everything prints correctly.
  library(P[i],character.only=TRUE)
}
rm(P)
rm(i)
```

## Data
Data for today's tutorial comes from three sources:
1) The data replication materials from Bricker, C., & LaCombe, S. (2021). The Ties that Bind Us: The Influence of Perceived State Similarity on Policy Diffusion. Political Research Quarterly, 74(2), 377-387. https://doi.org/10.1177/1065912920906611. 

The replication materials are from Dr. LaCombe's personal website, found here: https://www.scottlacombe.com/publications-and-working-papers.html

  In Bricker and LaCombe's paper, the authors test whether non-geographic characteristics of the states impact policy diffusion. Similarity is measured via a survey of citizen perceptions of their own states. Using the respondent scores, Bricker and LaCombe construct "similarity ties" between the states to test the impact of perceived state similarities in the policy innovation and diffusion.   
  
2) Data from the Census Bureau which defines the "Divisions" and "Regions" of states. "Divisions" is more granular definition of the region each state is classified under, and includes 9 named levels. "Regions" is a more broad set of categories and includes only 4 named levels. The data came from here: https://github.com/cphalpert/census-regions/blob/master/us%20census%20bureau%20regions%20and%20divisions.csv

3) Data on fatal accidents in each state from the fivethirtyeight github repository. This data includes a variable for the count of fatal accidents in each state in billions. The next columns describe the percentage of the count of accidents that were caused by a circumstance, such as speeding or drunk driving. Find the data here: https://github.com/fivethirtyeight/data/blob/master/bad-drivers/bad-drivers.csv

Today, we will test similarity of the states by their geographic similarity within regions (Division and Region). We will also test if bad driving is similar across states. There may be geographic features, laws, or social attitudes within each state that may cause a similar volume of fatal car accidents. 

```{r data}
#load Bricker and LaCombe's state network data
bl_data <- read_dta("dyadic_data.dta") 

#bad driver data 
driver <- read_csv("bad-drivers.csv")

#drop DC - it is not included in the state network data
driver <- driver[-c(9),] #from https://github.com/fivethirtyeight/data/blob/master/bad-drivers/bad-drivers.csv

#Census data
census <- read_csv("regions.csv") #from https://github.com/cphalpert/census-regions/blob/master/us%20census%20bureau%20regions%20and%20divisions.csv

#drop DC
census<- census[-c(8),] 

#Mapping package is picky; the capitalized "State" variable has to be lower case:
census$state <-census$State #attach lower case state variable to census data
driver$state <-driver$State  #attach lower case state variable to driver data
```
## Map
The argument of the paper is that measuring state similarity via citizen perceptions of their state is more complete than a geographic, contiguity measure. Here are the states visualized by a new geographic element, their Region, and subregion Division. Both variables come from the census data. 

A map for the density of fatal accidents across the states might also visualize whether geography impacts fatal driving, or if states with similar fatality counts are the same in a different way.
```{r map}
par(mfrow=c(1,2))
#A map of the states visualized by their Division identity
plot_usmap("states", labels= TRUE, data = census, values = "Division") + #This line says to plot the map with "states" outlined, to label each state, to use "census" data, and that the value to fill the states on is their "Division" identity
  scale_fill_brewer(blues9)   #the colors used to fill the states is from the ColorBrewer argument    

#A map of the states visualized by their Region identity
plot_usmap("states", labels= TRUE, data = census, values = "Region") + #This line says to plot the map with "states" outlined, to label each state, to use "census" data, and that the value to fill the states on is their "Region" identity
  scale_fill_brewer(blues9)  #the colors used to fill the states is from the ColorBrewer argument  

#The Denisty of the number of fatal car accidents in billions, by state
plot_usmap("states", labels= TRUE, data = driver, values = "num_fatal") + #This line says to plot the map with "states" outlined, to label each state, to use "driver" data, and that the value to fill the states on is their "num_fatal" value
  scale_fill_continuous(low = "white", high="#ce1b07", guide="none")  #the colors used to fill the states are from two chosen colors, which R will automatically scale to match continuous data. "num_fatal" is continuous, so the color argument is different from the discrete examples.  

```

## Network Object
To test measures of assortative mixing on the state network, the data must be prepared as a network object.
```{r netwob}
# add the states denoted as matrix sides (i &j) and their proportion of similarity
netvars <- bl_data[,c("state_i","state_j","proportion")]

# similarity is only measured in one year
# subset down to the 50x49 unique observations
unetvars <- unique(netvars)

# Similarity is measured as a percentage in the data. Here, the similarity threshold is set at 90%
ethresh <- quantile(unetvars$proportion,.9) #.9 dentoes a threshold of 90%
edges <- as.matrix(unetvars[unetvars$proportion > ethresh ,c(1,2)]) #next, a matrix of the edges is made based on the threshold argument made in the line above

# make a list of nodes
nodes <- sort(unique(c(as.matrix(unetvars[,c(1,2)])))) #creating a list of the nodes with the sort argument to sort by state name


# build the network object
state_net <- network.initialize(length(nodes)) #the network.initialize command from the network package creates a network based on the length of the nodes list 
network.vertex.names(state_net) <- nodes #this command sets the name of the vertex to the names of the nodes list 
state_net[edges] <- 1 # add the edge list made above to the network object

#Store Number of Fatal Death attribute
set.vertex.attribute(state_net,             # Store in state_net network object
                      "Division",            # What to name the attribute
                     census$Division)

#Store Number of Fatal Death attribute
set.vertex.attribute(x=state_net,             # Store in state_net network object
                     attrname = "num_fatal",            # What to name the attribute
                     driver$num_fatal)
```
## Visualize the Network
Division 
```{r Viz}
#Division Colors
div_colors <-ifelse(census$Division=="Pacific", "#CDFAFF", ifelse(census$Division=="West South Central", "#ce1b07", ifelse(census$Division=="East South Central", "#5684B0",
        ifelse(census$Division=="Mountain", "#e2803a", ifelse(census$Division=="New England", "#CEBB9D",ifelse(census$Division=="South Atlantic", "#C3D1E6",
        ifelse(census$Division=="West North Central", "#7A0200", ifelse(census$Division=="East North Central", "#CC5500", ifelse(census$Division=="Middle Atlantic", "#008080", "white")))))))))


set.seed(09022020)
xy_cool <- network.layout.fruchtermanreingold(state_net,layout.par = list(cool.exp=1)) #using FR coordinates 
plot(state_net, # network
     displaylabels=T, # display the labels
     coord = xy_cool,  # set the coordinates
     vertex.cex=3, # set the size of the vertices
     label.cex=0.5, # set the size of the labels 
     edge.col=rgb(150,150,150,100,maxColorValue=255),# edge colors
     label.pos=5, # put the labels in the middle
     vertex.col= div_colors) #use the node_colors list in the viz

```
Region
```{r Viz}
#Region Colors
reg_colors <-ifelse(census$Region=="South", "#CDFAFF", ifelse(census$Region=="West", "#ce1b07", ifelse(census$Region=="Midwest", "#5684B0", "white")))


set.seed(09022020)
xy_cool <- network.layout.fruchtermanreingold(state_net,layout.par = list(cool.exp=1)) #using FR coordinates 
plot(state_net, # network
     displaylabels=T, # display the labels
     coord = xy_cool,  # set the coordinates
     vertex.cex=3, # set the size of the vertices
     label.cex=0.5, # set the size of the labels 
     edge.col=rgb(150,150,150,100,maxColorValue=255),# edge colors
     label.pos=5, # put the labels in the middle
     vertex.col= reg_colors) #use the node_colors list in the viz

```
Fatal Driving Accidents 
```{r Viz}
#Division Colors
fatal_colors <-ifelse(driver$num_fatal>=18, "#7A0200", ifelse(driver$num_fatal>=15 , "#ce1b07", ifelse(driver$num_fatal>=10 , "#CEBB9D", "white")))


set.seed(09022020)
xy_cool <- network.layout.fruchtermanreingold(state_net,layout.par = list(cool.exp=1)) #using FR coordinates 
plot(state_net, # network
     displaylabels=T, # display the labels
     coord = xy_cool,  # set the coordinates
     vertex.cex=3, # set the size of the vertices
     label.cex=0.5, # set the size of the labels 
     edge.col=rgb(150,150,150,100,maxColorValue=255),# edge colors
     label.pos=5, # put the labels in the middle
     vertex.col= fatal_colors) #use the node_colors list in the viz
```
## Assortative Mixing: Non-Weighted Graph
num_fatal=0.11, low homophily
fatal_no_provisions= -0.04, low heterophily
Division=0.41, mild homophily
Region= 0.74, high homophily
```{r assortmix}
#Assortatative Mixing 
#Continuous Examples 
assortment.continuous(state_net,driver$num_fatal,weighted=FALSE) #the assortnet continuous command on the num_fatal variable. Note you can call the variable from the vector of a dataset-you do not need to assign it as a network attribute. 
assortment.continuous(state_net,driver$fatal_no_previous,weighted=FALSE)  #the assortnet continuous command on the fatal_no_previous variable.

#Discrete Examples
assortment.discrete(state_net,census$Division,weighted=FALSE)  #the assortnet discrete command on the Division variable.
assortment.discrete(state_net,census$Region,weighted=FALSE)  #the assortnet discrete command on the Region variable.

```
## Weighted Assortatative Mixing 
Weighting the graph impacts the assortatative mixing results:
num_fatal=0.09, low homophily (lower than unweighted ex.)
fatal_no_provisions= -0.04, low heterophily (same as unweighted ex.)
Division=0.16, low homophily (lower than unweighted ex.)
Region= 0.35, mild homophily (lower than unweighted ex.)
```{r assortweight}
#Assortatative Mixing 
#Continuous Examples
state_netW <- state_net[,] #make a new network object from state_net
state_netW[as.matrix(unetvars[,c(1,2)])] <- unetvars$proportion #add weights to the network using the proportion variable
assortment.continuous(state_netW,driver$num_fatal,weighted=T) #the same code from the non-weighted example on num_fatal, except now weighted=T
assortment.continuous(state_netW,driver$fatal_no_previous,weighted=T) #the same code from the non-weighted example on fatal_no_previous, except now weighted= T

#Discrete Examples
assortment.discrete(state_netW, census$Division,weighted=T) #the same code from the non-weighted example on Division, exceptt now weighted=T
assortment.discrete(state_netW, census$Region,weighted=T) #the same code from the non-weighted example on Region, except now weighted=T
```
## igraph example
The igraph package has more assortative mixing commands. However, you will need to make your network object using igraph. Additionally, any commands from the network package should be ran before loading igraph, as the two packages are clunky together. 

There is a discrete and continuous example for the igraph package, just as in the assortnet package. Note that the values are the same as the unweighted network examples in assortnet.

Additonally, igraph allows you to test the assortativity degree. Here, we are testing whether nodes with high perceived state similarity scores frequently connect together. It is essentially a Pearson's correlation measure where we ask if similar nodes are correlated with each other. 
```{r igraph}
library(igraph) #load igraph
#install.packages("intergraph") uncomment and run this line if you do not have the intergraph pacakge installed
library(intergraph) #load intergraph 

grph <- asIgraph(state_net) #make the igraph network object from state_net

assortativity(grph, types1 = driver$num_fatal, types2 = NULL, directed = TRUE) # assortativity commannd is for continuous variables, such as num_fatal
assortativity_nominal(grph, types = as.factor(census$Region), directed = TRUE) # assortativity_nominal command is for discrete variable such as Region
assortativity_degree(grph, directed = TRUE) #This command tests the degree of assortativity on nodes of the graph 

```

## Pacakge Info
assortnet: https://cran.r-project.org/web/packages/assortnet/assortnet.pdf
igraph: https://igraph.org/r/html/1.3.2/assortativity.html 
